---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r loadingData, echo=TRUE}
data <- read.csv("activity.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)
```


## What is mean total number of steps taken per day?
1. Calculate the total number of steps taken per day.
```{r , echo=TRUE}
total_steps_per_day <- aggregate(x = data$steps, by = list(data$date), FUN = sum, 
                                 na.rm = TRUE)
names(total_steps_per_day) <- c("date", "total_steps")
```

2. Make a histogram of the total number of steps taken each day.
```{r histogram, fig.width=10, echo=TRUE}
library(ggplot2)
q <- qplot(y = total_steps, x = date, data = total_steps_per_day, geom = "histogram", 
      stat = "identity", xlab = "", ylab = "Total number of steps",
      main = "Total Number of Steps taken each day")
q + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

3. Calculate the mean and median of the total number of steps taken each day.
```{r, echo=TRUE}
mean(total_steps_per_day$total_steps, na.rm = TRUE)
```

```{r, echo=TRUE}
median(total_steps_per_day$total_steps, na.rm = TRUE)
```


## What is the average daily activity pattern?
1. Make a time series plot of the 5-minute interval (x-axis) and the average 
number of steps taken, averaged across all days (y-axis)
```{r timeseriesplot, echo=TRUE}
interval_average_steps <- aggregate(x = data$steps, by = list(data$interval), data = data, 
                                    FUN = mean, na.rm = TRUE)
names(interval_average_steps) <- c("interval", "avg_steps")
plot(x = interval_average_steps$interval, y = interval_average_steps$avg_steps, 
     type = "l", xlab = "Interval", ylab = "Average steps")
```

2. Which 5-minute interval, on average across all the days in the dataset, 
contains the maximum number of steps?
```{r , echo=TRUE}
interval_average_steps[which.max(interval_average_steps$avg_steps), c("interval")]
```


## Imputing missing values
1. Calculate and report the total number of missing values in the dataset.
```{r, echo=TRUE}
sum(is.na(data))
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in. (In this, I will use the average 5-minute interval)
```{r, echo=TRUE}
merged_data <- merge(x = data, y = interval_average_steps, by = "interval", all.x = TRUE)
merged_data[is.na(merged_data$steps), c("steps")] <- 
    merged_data[is.na(merged_data$steps), c("avg_steps")]
```

4. Make a histogram of the total number of steps taken each day and Calculate and 
report the mean and median total number of steps taken per day.
```{r, echo=TRUE}
total_steps_with_filled_NAs_each_day <- aggregate(x = merged_data$steps, 
                                                  by = list(merged_data$date),
                                                  FUN = sum, na.rm = TRUE)
names(total_steps_with_filled_NAs_each_day) <- c("date", "avg_steps")
q <- qplot(y = avg_steps, x = date, data = total_steps_with_filled_NAs_each_day, 
           geom = "histogram", stat = "identity", xlab = "", 
           ylab = "Total number of steps", main = "Total Number of Steps taken each day")
q + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r, echo=TRUE}
mean(total_steps_with_filled_NAs_each_day$avg_steps, na.rm = TRUE)
```

```{r, echo=TRUE}
median(total_steps_with_filled_NAs_each_day$avg_steps, na.rm = TRUE)
```


## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r, echo=TRUE}
merged_data$date <- as.Date(x = merged_data$date)
Sys.setlocale("LC_TIME", "C") # set to english
merged_data$weekday <- as.factor(ifelse(weekdays(merged_data$date) %in% 
                                            c("Saturday", "Sunday"), "Weekend", "Weekday"))
```

2. Make a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).
```{r, echo=TRUE}
average_steps_weekday <- aggregate(x = merged_data$steps, 
                                   by = list(merged_data$interval, merged_data$weekday),
                                   FUN = mean, na.rm = TRUE)
names(average_steps_weekday) <- c("interval", "weekday", "avg_steps")
plot_weekdays <- qplot(y = avg_steps, x = interval, data = average_steps_weekday, 
                       geom = "line", xlab = "Interval", ylab = "Average number of steps",
                       facets = . ~ weekday, main = "Time series plot of interval and average steps taken across Weekdays/Weekends")
plot_weekdays
```

